---
title: "P8105 Homework 2"
author: "Xiaobo Shen"
date: "2026-10-07"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
```

## Problem 1
```{r}
subway_df =
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(
      Route8 = col_character(),
      Route9 = col_character(),
      Route10 = col_character(),
      Route11 = col_character()
    )
  )

head(subway_df)
```
```{r}
subway_df =
  subway_df |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    route1:route11, entry, vending, entrance_type, ada
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )

dim(subway_df)
head(subway_df)
```
The cleaned dataset contains 1,868 rows and 19 columns.
The retained variables are: subway line, station name, station latitude/ longitude, routes, entry, vending, entrance type, and ADA.

I cleaned the column names and selected the required variables, then converted entry from YES/NO to TRUE/FALSE. 

The data are not tidy yet, because route1 through route11 are spread across multiple columns.

```{r}
subway_df |>
  distinct(line, station_name) |>
  nrow()
```
There are 465 distinct stations.

```{r}
subway_df |>
  filter(ada == TRUE) |> 
  distinct(line, station_name) |>
  nrow()
```
84 stations are ADA compliant.

```{r}
subway_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean()
```
About 37.7049% station entrances / exits without vending allow entrance.

```{r}
subway_tidy =
  subway_df |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  ) |>
  mutate(route_number = as.integer(route_number)) |>
  filter(!is.na(route_name))

head(subway_tidy)
```

```{r}
subway_tidy |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()
```

There are 60 distinct stations that serve the A train.

```{r}
subway_tidy |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
Of the stations that serve the A train, 17 are ADA compliant.

## Problem 2

## Problem 3