---
title: "P8105 Homework 2"
author: "Xiaobo Shen"
date: "2026-10-07"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
```

## Problem 1
```{r}
subway_df =
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(
      Route8 = col_character(),
      Route9 = col_character(),
      Route10 = col_character(),
      Route11 = col_character()
    )
  )

head(subway_df)
```
```{r}
subway_df =
  subway_df |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    route1:route11, entry, vending, entrance_type, ada
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )

dim(subway_df)
head(subway_df)
```
The cleaned dataset contains 1,868 rows and 19 columns.
The retained variables are: subway line, station name, station latitude/ longitude, routes, entry, vending, entrance type, and ADA.

I cleaned the column names and selected the required variables, then converted entry from YES/NO to TRUE/FALSE. 

The data are not tidy yet, because route1 through route11 are spread across multiple columns.

```{r}
subway_df |>
  distinct(line, station_name) |>
  nrow()
```
There are 465 distinct stations.

```{r}
subway_df |>
  filter(ada == TRUE) |> 
  distinct(line, station_name) |>
  nrow()
```
84 stations are ADA compliant.

```{r}
subway_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean()
```
About 37.7049% station entrances / exits without vending allow entrance.

```{r}
subway_tidy =
  subway_df |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  ) |>
  mutate(route_number = as.integer(route_number)) |>
  filter(!is.na(route_name))

head(subway_tidy)
```

```{r}
subway_tidy |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()
```

There are 60 distinct stations that serve the A train.

```{r}
subway_tidy |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
Of the stations that serve the A train, 17 are ADA compliant.

## Problem 2

```{r}
mr_trash =
  readxl::read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N742"
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    year = as.integer(year),
    wheel = "Mr. Trash Wheel"
  )

dim(mr_trash)
head(mr_trash)
```

```{r}
professor_trash =
  readxl::read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    na = "dive",  ## Treat the non-numeric bottle count as missing
    range = "A2:M139"
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    wheel = "Professor Trash Wheel"
  )

dim(professor_trash)
head(professor_trash)
```
```{r}
gwynnda_trash =
  readxl::read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L394"
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    wheel = "Gwynnda"
  )

dim(gwynnda_trash)
head(gwynnda_trash)
```
```{r combine}
combine_trash =
  bind_rows(mr_trash, professor_trash, gwynnda_trash)

dim(combine_trash)
head(combine_trash)
```

The combined dataset contains 1,269 observations and 15 variables, from Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. Each observation represents one dumpster collection. Key variables are collection date, trash weight in tons, trash volume in cubic yards, and counts of plastic bottles and cigarette butts.

```{r}
combine_trash |>
  filter(wheel == "Professor Trash Wheel") |>
  pull(weight_tons) |>
  sum(na.rm = TRUE)
```
The total weight of trash collected by Professor Trash Wheel is 294.96 tons based on the available data.

```{r}
combine_trash |>
  filter(
    wheel == "Gwynnda",
    year == 2022,
    month == "June"
  ) |>
  pull(cigarette_butts) |>
  sum(na.rm = TRUE)
```

the total number of cigarette butts collected by Gwynnda is 18120

## Problem 3